Towards Practical First-Order Model Counting

### 2024-04-18 doc/paper.tex
Section 3.2

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item say something about what $\expr{}$ can contain (function calls,
          various arithmetic operations, summations, non-negative integer
          constants, variables introduces by the function signature or
          summations)
  \end{itemize}
}

@@

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Crane. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we get the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, giving us $g(l, 0) = 1$.
\end{example}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item \Cref{alg2} now supports any base case domain sizes
    \item NOTE: if $C' = \emptyset$, then the clause becomes irrelevant and
          doesn't need smoothing, so can be skipped
  \end{itemize}
}

@@

\todo[inline,caption={}]{ Turn these notes into a coherent description (that
  doesn't mention \textsc{Crane}):
  \begin{itemize}
    \item Each base case identified by \FindBaseCases can be interpreted as
          assuming some domain...
    \item \Propagate (\cref{alg2}) to find the transformed formula corresponding
          to each base case identified by \FindBaseCases and call \textsc{Crane}
          on the formula to obtain the required base cases.
    \item Clauses where all literals have variables quantified over the empty
          domain are still removed.
    \item $C'$ is guaranteed to be non-empty.
    \item \Cref{line:smoothing1,line:smoothing2} are explained in
          \cref{sec:smoothingbase}.
    \item For a literal or a clause $C$, let $\Doms(C)$ denote the set of the
          corresponding domains. Make sure an example illustrates this.
  \end{itemize}
}
\todo[inline]{Maybe prove consistency: if two base case overlap, their numerical
  answers also overlap.}

@@

\todo[inline]{Can I prove that the additions to smoothing outlined in the rest
  of this section `do the right thing'?}

@@

\todo[inline]{Can I use an example based on permutations or something similar?}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Maybe add some complete examples of C++ programs in the supplementary
          material.
          \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item Full stop at the end of a single-sentence caption?
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
          \end{itemize}
  \end{itemize}
}

>>>>

\crefalias{enumi}{type}
\crefname{type}{Type}{Types}
\creflabelformat{type}{#2\textup{#1}#3}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item say something about what $\expr{}$ can contain (function calls,
          various arithmetic operations, summations, non-negative integer
          constants, variables introduces by the function signature or
          summations)
    \item The definition of a function call contradicts the base case finding
          algorithm. Maybe rephrase it to only look at function calls from $f$
          to itself and assume:
          \begin{itemize}
            \item no mutual recursion,
            \item these calls in particular have this form
          \end{itemize}
  \end{itemize}
}

@@

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Crane. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we get the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, giving us $g(l, 0) = 1$.

  Note that these base cases overlap when $l = m = 0$ but are consistent with
  each other since $0^{0} = 1$. More generally, let $\phi$ be a formula with two
  domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the
  model count of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is
  the same as the model count of \Propagate{$\phi$, $\Gamma$, $m$} assuming
  $|\Delta| = n$.
\end{example}

\todo[inline]{This last claim sounds like something that should be proven.}

@@

@@

\Propagate (\cref{alg2}) modifies formula $\phi$ with the assumption that domain
$\Delta$ has size $n \in \mathbb{N}_{0}$. In the case of $n=0$, many clauses
become vacuously satisfied and can be removed. In the case of
$n > 0$\footnote{None of the formulas considered in this work had $n > 1$.}, we
perform partial grounding, using constants to replace all variables quantified
over $\Delta$. \Cref{alg2} considers these two cases separately. For a literal
or a clause $C$, we write $\Doms(C)$ to denote the set of corresponding domains.

In the case of $n = 0$, consider three types of clauses:
\begin{enumerate*}[label=(\roman*)]
  \item those that do not mention $\Delta$,\label{type1}
  \item those in which every literal contains variables quantified over
  $\Delta$, and\label{type2}
  \item those that have some literals with variables quantified over $\Delta$
  and some without.\label{type3}
\end{enumerate*}
\Cref{type1} clauses are transferred to the new formula $\phi'$ unchanged. For
\cref{type2} clauses, $C' = \emptyset$, so these clauses are filtered out. One
might think that the same should be done with \cref{type3} clauses, however,
\cref{line:smoothing1,line:smoothing2} perform a new kind of smoothing, the
explanation of which we defer to \cref{sec:smoothingbase}.

In the case of $n>0$, we introduce $n$ new constants. Consider an arbitrary
clause $C \in \phi$ and let $m \in \mathbb{N}_{0}$ be the number of variables in
$C$ quantified over $\Delta$. If $m=0$, then, similarly to the previous case, we
add $C$ directly to $\phi'$. Otherwise, we add a clause to $\phi'$ with every
possible way of replacing the $m$ variables in $C$ with some combination of the
$n$ new constants.

\begin{example}
  Let us consider the clause $C \equiv \forall x \in \Gamma\text{.
  }\forall y, z \in \Delta\text{. } \neg P(x, y) \lor \neg P(x, z) \lor y=z$.
  Then
  $\Doms(C) = \Doms(\neg P(x, y)) = \Doms(\neg P(x, z)) = \{\, \Gamma, \Delta \,\}$,
  and $\Doms(y=z) = \{\, \Delta \,\}$. A call to \Propagate{$\{\, C \,\}$,
    $\Delta$, $3$} would produce the following formula with nine clauses:
  \begin{align*}
    \forall x \in \Gamma\text{. }\neg P(x, c_{1}) \lor& \neg P(x, c_{1}) \lor c_{1}=c_{1}\\
    \forall x \in \Gamma\text{. }\neg P(x, c_{1}) \lor& \neg P(x, c_{2}) \lor c_{1}=c_{2}\\
    \vdots&\\
    \forall x \in \Gamma\text{. }\neg P(x, c_{3}) \lor& \neg P(x, c_{3}) \lor c_{3}=c_{3},\\
  \end{align*}
  where $c_{1}$, $c_{2}$, and $c_{3}$ are the new constants.
\end{example}

@@

\todo[inline,caption={}]{
\begin{itemize}
  \item Cite something when introducing the idea of smoothing. Maybe both
        propositional and first-order sources.
  \item Can I prove that the additions to smoothing outlined in the rest of this
        section `do the right thing'?
\end{itemize}
}

@@

\todo[inline]{Must include an example of propagating the empty-domain
  assumption (ideally based on previous examples).}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Maybe add some complete examples of C++ programs in the supplementary
          material.
          \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item Full stop at the end of a single-sentence caption?
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
            \item Something to think about that could inspire some kind of
                  theorems: how can I assume that a domain is empty if there are
                  constants associated with it?
          \end{itemize}
  \end{itemize}
}
